Quantum H theorem, variational entropy, and relaxation processes

### 2021-01-25 tex/classical.tex
Additions to intro.

The distribution functions, $\{f'_M(\vec{v},t)\}$, depend implicitly on the position of the cells,
relative to the global system, and on the velocity
$\vec{v}$ and time $t$. Notice that each $f'_M$ can be formally extended to the complete coordinate space
by defining each $f'_M$ to be zero outside the $M$-th cell, in such a manner that
the distribution function of the complete system is a piece-wise sum
of $\{f'_M(\vec v,t)\}$:
%
\begin{equation}
   f'(\vec r,\vec v,t)=\sum_{M=1}^Kf'(\vec r_M,\vec v,t)=\sum_{M=1}^Kf'_M(\vec v,t).
\end{equation}
%
Here $\vec r_M$ is the center of the cell of index $M$, and $f'_M(\vec v,t)\neq f'_N(\vec v,t)$
for $M\neq N$. This extended definition
allows us to omit the symbol $\delta V_M$ on all integrals performed over the cell volume.
In terms of $f'_M(\vec{v},t)$ and a local variable of energy, $\epsilon(\vec{v})$, we have:
%
\begin{subequations}\label{eq:cellrestrictions}
\begin{eqnarray}
    \mathcal{H}_M' & = &  \int f'_M(\vec{v},t) \ln f'_{M}(\vec{v},t)
      \mathrm{d}^{3}v \label{eq:Hcell},\\
    \mathcal{N}_M' & = & \int f'_{M}(\vec{v} ,t) \mathrm{d}^{3}v,\label{eq:Ncell}\\
    \mathcal{E}_M' & = & \int f'_{M}(\vec{v},t)\epsilon(\vec{v}) \mathrm{d}^{3}v\label{eq:Ecell}.
\end{eqnarray}
\end{subequations}
%
Notice that assuming that every $f'_M$ is homogeneous implies that we are accepting the validity of the
local equilibrium hypothesis. In addition, the set $\{f'_{M}(\vec{v},t)\}$ must satisfy the following restrictions:
%
\begin{subequations}\label{eq:micro}
\begin{equation}\label{eq:micron}
    \sum_{M=1}^{K}\int f'_M(\vec{v},t)\mathrm{d}^3v =N
\end{equation}
and
\begin{equation}\label{eq:microe}
    \sum_{M=1}^{K}\int f'_M(\vec{v},t)\epsilon(\vec{v})\mathrm{d}^3v=E.
\end{equation}
\end{subequations}
%

>>>>

The distribution functions, $\{f'_M(\vec{v},t)\}$, depend implicitly on the position of the cells,
relative to the global system, and on the velocity
$\vec{v}$ and time $t$. Notice that each $f'_M$ can be formally extended to the complete coordinate space
by defining each $f'_M$ to be zero outside the $M$-th cell, in such a manner that
the distribution function of the complete system is a piece-wise sum
of $\{f'_M(\vec v,t)\}$:
%
\begin{equation}
   f'(\vec r,\vec v,t)=\sum_{M=1}^Kf'(\vec r_M,\vec v,t)=\sum_{M=1}^Kf'_M(\vec v,t).
\end{equation}
%
Here, $\vec r_M$ is the center of the cell of index $M$, and $f'_M(\vec v,t)\neq f'_N(\vec v,t)$
for $M\neq N$. This extended definition
allows us to omit the symbol $\delta V_M$ on all integrals performed over the cell volume.
In terms of $f'_M(\vec{v},t)$ and a local variable of energy, $\epsilon(\vec{v})$, we have:
%
\begin{subequations}\label{eq:cellrestrictions}
\begin{eqnarray}
    \mathcal{H}_M' & = &  \int f'_M(\vec{v},t) \ln f'_{M}(\vec{v},t)
      \mathrm{d}^{3}v \label{eq:Hcell},\\
    \mathcal{N}_M' & = & \int f'_{M}(\vec{v} ,t) \mathrm{d}^{3}v,\label{eq:Ncell}\\
    \mathcal{E}_M' & = & \int f'_{M}(\vec{v},t)\epsilon(\vec{v}) \mathrm{d}^{3}v\label{eq:Ecell}.
\end{eqnarray}
\end{subequations}
%
Notice that assuming every $f'_M$ to be homogeneous implies that we are accepting the validity of the
local equilibrium hypothesis. In addition, the set $\{f'_{M}(\vec{v},t)\}$ must satisfy the following restrictions:
%
\begin{subequations}\label{eq:micro}
\begin{equation}\label{eq:micron}
    \sum_{M=1}^{K}\int f'_M(\vec{v},t)\mathrm{d}^3v =N
\end{equation}
and
\begin{equation}\label{eq:microe}
    \sum_{M=1}^{K}\int f'_M(\vec{v},t)\epsilon(\vec{v})\mathrm{d}^3v=E.
\end{equation}
\end{subequations}
%

### 2021-01-25 tex/introduction.tex
Additions to intro.

The theoretical basis and the procedures that allow to describe classical equilibrium systems
are well-stablished. These procedures can be applied to treat a wide range of the systems
present in nature, and include both the macroscopic phenomenological methods (thermodynamics)
as well as the microscopic description (statistical mechanics). For instance,
in the Kinetic Theory of gases, the behavior of a dilute classical gas
is described through the Boltzmann transport equation \cite{bib:huang},
and the time-evolution of a system towards the equilibrium is proved by
means of the Boltzmann $H$-theorem, wherein the
spatial-homogeneous distribution function hypothesis plays a crucial role.

In contrast, for quantum equilibrium systems, the construction of a framework,
with the same level of
success and universality as the classical version, is still an open problem. For instance,
in order to obtain a complete correspondence principle, between classical mechanics
and quantum mechanics, the quantum analogous of both the Boltzmann $H$-theorem
and the Boltzmann transport equation have not attained a fully accepted form.
In this context, Tolman was one of the earliest physicists to propose a quantum
$H$-theorem \cite{bib:tolman}, using probability transition relations, the random
phases hypothesis, and a spatially homogeneous distribution function.
Also, Tolman proposed a potential quantum analogous of the transport equation,
in terms of the occupation number,
by applying time perturbation theory. Further attempts, under the quantum operator
formalism, have addressed the description of quantum transport phenomena by means of the Hamiltonian
of the system and the master equation (which here is the analogous
of the Boltzmann transport equation)
\cite{bib:grabert1974,bib:wang2014,%
bib:angel2017,
bib:amato2020,bib:nicacio2015,bib:hussein2014}. However, these approaches 
are not consistent with the classical-quantum correspondence principle.
In the same line of thought, some authors have proposed $H$-functionals
and attempted to proof a
quantum $H$-theorem \cite{bib:gorban2014,bib:bennaim2017,bib:silva2010,%
bib:deroeck2006,bib:acharya2019,bib:kastner2017,bib:han2015,bib:das2018,bib:vonneumann2010}.
However, whether or not the homogeneous distribution function hypothesis is assumed, or
if their framework obeys the correspondence principle is unclear or not discussed whatsoever.
Furthermore, the general validity
of the quantum $H$-theorem, the second law of thermodynamics, and the
interpretation of the quantum entropy is still in discussion %[19-26].
\cite{bib:silva2010,bib:acharya2019,bib:kastner2017,bib:han2015,bib:brown2008,%
bib:dragoljub2009,bib:vonneumann2010,%
bib:syros1999,bib:lesovik2016,
bib:lesovik2019}.

@@

The organization of this article is as follows.
In Section~\ref{sec:classical}, we highlight, to our purposes, the most
fundamental assumptions required to proof the Boltzmann $H$-theorem,
we provide an alternative method to obtain the Maxwell-Boltzmann
distribution using the variational method, and the demonstration of
an alternative $H$-theorem for classical systems. In Section~\ref{sec:quantum}
we review the Tolman proposal for the quantum
version of the $H$-theorem (quantum $H$-theorem)
\commr{and how the Bose-Einstein and Fermi-Dirac distributions are treated within
this framework},
and our alternative proof of the quantum $H$-theorem. In 
Section~\ref{sec:qccorrespondence}
the classical-quantum correspondence between the
quantum $H$-functional and the classical $H$-functional is analyzed.
In Section~\ref{sec:relaxproc} we explore how relaxation processes occur
in a quantum ideal gas; here we provide a time-evolution equation that could be
considered the analogous of the classical Boltzmann transport equation. Finally,
we discuss some key ideas derived from our approach and close with a brief summary
in Section~\ref{sec:disscussion}.

>>>>

The theoretical basis and the procedures that allow to describe equilibrium systems
are well-stablished. These procedures can be applied to treat a wide range of the systems
present in nature, and include both the macroscopic phenomenological methods (thermodynamics)
as well as the microscopic description (statistical mechanics).
(Of course, regarding out-of equilibrium systems, the present situation is still a challenge.) 
For instance,
in the Kinetic Theory of gases, the behavior of a dilute classical gas
is described through the Boltzmann transport equation \cite{bib:huang},
and the time-evolution of a system towards the equilibrium is proved by
means of the Boltzmann $H$-theorem, wherein the
spatial-homogeneous distribution function hypothesis plays a crucial role.

On the other hand, for quantum equilibrium systems, the construction of a framework,
with the same level of
success and universality as the classical version, is still an open problem. For instance,
in order to obtain a complete correspondence principle, between classical mechanics
and quantum mechanics, the quantum analogous of both the Boltzmann $H$-theorem
and the Boltzmann transport equation have not attained a fully accepted form.
In this context, Tolman was one of the earliest physicists to propose a quantum
$H$-theorem \cite{bib:tolman}, using probability transition relations, the random
phases hypothesis, and a spatially homogeneous distribution function.
Also, Tolman proposed a potential quantum analogous of the transport equation,
in terms of the occupation number,
by applying time perturbation theory. Further attempts, under the quantum operator
formalism, have addressed the description of quantum transport phenomena by means of the Hamiltonian
of the system and the master equation (which here is the analogous
of the Boltzmann transport equation)
\cite{bib:grabert1974,bib:wang2014,%
bib:angel2017,
bib:amato2020,bib:nicacio2015,bib:hussein2014}. However, these approaches 
are not consistent with the classical-quantum correspondence principle.
In the same line of thought, some authors have proposed $H$-functionals
and attempted to proof a
quantum $H$-theorem \cite{bib:gorban2014,bib:bennaim2017,bib:silva2010,%
bib:deroeck2006,bib:acharya2019,bib:kastner2017,bib:han2015,bib:das2018,bib:vonneumann2010}.
However, whether or not the homogeneous distribution function hypothesis is assumed, or
if their framework obeys the correspondence principle is unclear or not discussed whatsoever.
Furthermore, the general validity
of the quantum $H$-theorem, the second law of thermodynamics, and the
interpretation of the quantum entropy is still in discussion %[19-26].
\cite{bib:silva2010,bib:acharya2019,bib:kastner2017,bib:han2015,bib:brown2008,%
bib:dragoljub2009,bib:vonneumann2010,%
bib:syros1999,bib:lesovik2016,
bib:lesovik2019}.

@@

The organization of this article is as follows.
In Section~\ref{sec:classical}, we highlight, to our purposes, the most
fundamental assumptions required to proof the Boltzmann $H$-theorem,
we provide an alternative method to obtain the Maxwell-Boltzmann
distribution using the variational method, followed by our proposal of
an alternative $H$-functional for classical systems, and the demonstration of
the respective $H$-theorem . In Section~\ref{sec:quantum}
we review the Tolman proposal for the quantum
version of the $H$-theorem (quantum $H$-theorem)
and how the Bose-Einstein and Fermi-Dirac distributions are treated within
this framework, subsequently we present our proposal for a quantum $H$-functional
and the proof of the corresponding quantum $H$-theorem. In 
Section~\ref{sec:qccorrespondence}
the classical-quantum correspondence between the
quantum and classical $H$-functionals is analyzed.
In Section~\ref{sec:relaxproc} we explore how relaxation processes occur
in a quantum ideal gas; \commb{here we provide a time-evolution equation that could be
considered the analogous of the classical Boltzmann transport equation.} Finally,
we discuss some key ideas derived from our approach and close with a brief summary
in Section~\ref{sec:disscussion}.
